---
title: "User Guide: Implementing an Elo Rating System in R"
author: "Max Buettenback"
format: html
bibliography: references.bib
---

<<<<<<< HEAD
# Introduction {.unnumbered}

Elo ratings are a popular way for ranking competitors in head-to-head competitions. Each competitor has a certain rating that updates after each and every match depending on the outcome. Rating differences can be used to predict future match results too. This guide explains how to implement an Elo rating system in R using the `PlayerRatings` package. It also includes instructions for loading and preparing match data, calculating Elo ratings, visualizing ratings and interpreting the results. The workflow is based on the concepts described by Glickman and Jones.[@glickman2024]

# Prerequisites {.unnumbered}

Before beginning this guide, users should have:

- R and RStudio installed  
- Basic familiarity with R (loading data and using packages)  
- A dataset containing `player1`, `player2`, and `result` columns  
- Match data arranged in chronological order  

Ensuring these requirements are met will allow the Elo rating system to function correctly.

# Step 1: Data Preparation {.unnumbered}

Before model implementation, the dataset must be cleaned and formatted properly. Player names should be consistent across all observations, and matches must be ordered chronologically to ensure accurate rating updates.

```{r message=FALSE}
library(dplyr)

# Placeholder dataset for demonstration
elo_data <- data.frame(
  player1 = c("Player A", "Player B", "Player A", "Player C"),
  player2 = c("Player B", "Player C", "Player C", "Player D"),
  result  = c("player1", "player2", "player1", "player2")
)

# Encode results numerically for Elo
elo_data <- elo_data %>%
  mutate(score1 = ifelse(result == "player1", 1, 0),
         score2 = ifelse(result == "player2", 1, 0)) %>%
  select(player1, player2, score1, score2)

elo_data
```

::: callout-note
The dataset above is a placeholder created directly in R for demonstration purposes. In practice, users should replace it with their own match data containing the columns player1, player2, and result.
:::

# Step 2: Loading Required Package {.unnumbered}

The Elo rating system is implemented using the `PlayerRatings` package [@playerratings]. The `PlayerRatings` package provides functions to help dynamically update player ratings
```
library(PlayerRatings)
```
# Step 3: Parameter Specification {.unnumbered}

The K-factor controls how much a player's rating changes after each match. The higher the K-factor means the faster the ratings change. A lower K-factor makes ratings a lot more stable.

The choice of the K-factor number depends a lot on the amount of data available. If there is aren't a lot of matches or the player is new then having a higher K-factor, like 32, is useful because it allows ratings to adjust quickly. Then on the other hand, if there is a more established player with a lot more matches, a lower K-factor, like 16-24, is better because ratings don't change as quickly.
```
k_factor <- 32
```
# Step 4: Model Estimation or Updating {.unnumbered}

The Elo model updates ratings after each match by comparing what actually happened to what was expected. If a player performs better then expected it goes up and if not it goes down.

This process happens one match at a time, so ratings change as more data is added. Because of this, the matches must be in chronological order to get correct results.
=======
## Introduction {.unnumbered}

Elo ratings are a popular way for ranking competitors in head-to-head competitions. Each competitor has a certain rating that updates after each and every match depending on the outcome. Rating differences can be used to predict future match results too. This guide explains how to implement an Elo rating system in R using the "PlayerRatings" package. It also includes instructions for loading and preparing match data, calculating Elo ratings, visualizing ratings and interpreting the results. The workflow is based on the concepts described by Glickman and Jones.[@glickman2024]

Some things that users should have before starting is R and Rstudio installed and they should also be familiar with basic R operations such as loading datasets and using packages. The dataset should also include the following columns: player1, player2, and results. Matches should be in chronological order, and any results that are missing should be removed to make sure everything is accurate.

## Step 1: Data Preparation {.unnumbered}

Before running any model, the dataset must be cleaned and formatted. First, check that all competitor names are consistent and correct any typos. Ensure matches are in chronological order, which is particularly important for dynamic ratings like Elo. Next, encode match outcomes numerically: wins as 1 and losses as 0. (For the sake of the user guide I will have a fake placeholder csv called matches).
```
library(dplyr)

matches <- read.csv("matches.csv")  # Replace with your file path

elo_data <- matches %>%
  mutate(score1 = ifelse(result == "player1", 1, 0),
         score2 = ifelse(result == "player2", 1, 0)) %>%
  select(player1, player2, score1, score2)

head(elo_data)
```
## Step 2: Model Selection {.unnumbered}

Next we decide which rating model is appropriate for our data. For static models, like Bradley–Terry, ratings assume outcomes are independent and do not change over time. For dynamic models, like Elo or Glicko, ratings update after each match. In this guide, we use the Elo model for dynamic updating.
```
library(PlayerRatings)
```
## Step 3: Parameter Specification {.unnumbered}

Now we select starting ratings for each player and choose tuning parameters. In Elo, the K-factor determines how much ratings change after each match. A higher K-factor makes ratings more sensitive to match results.
```
k_factor <- 32
```
## Step 4: Model Estimation or Updating {.unnumbered}

After that we use the Elo model to calculate or update ratings based on match outcomes. Each match updates player ratings using the specified K-factor.
>>>>>>> 1e05f80108bfce3289b835ae9a0ad58ca865960f
```
elo_ratings <- elo(elo_data, k = k_factor)
elo_ratings
```
<<<<<<< HEAD
# Step 5: Evaluation {.unnumbered}

After calculating the ratings, it is useful to evaluate and visualize the results. This helps identify which players are ranked higher and whether the ratings appear reasonable.

The figure below shows an example of Elo ratings for a set of players.

```{r fig-example-ratings}
#| fig-cap: "Example Elo ratings for players"
#| message: false

players <- c("Player A", "Player B", "Player C", "Player D")
ratings <- c(1600, 1500, 1450, 1300)

barplot(ratings,
        names.arg = players,
        xlab = "Players",
        ylab = "Elo Rating",
        main = "Example Elo Ratings")
```

As shown in @fig-example-ratings, players with higher ratings are considered stronger. These visualizations can help check that the model results make sense and are easy to interpret.

# Step 6: Interpretation {.unnumbered}

After calculating the Elo ratings, you can use them to tank players and also estimate changes of winning future matches.

- Players with the **same rating** have about a 50% chance of winning.  
- A player with **200 points higher** than their opponent would win about 75% of the time.  
- A player with **400 points higher** than their opponent would win about 90% of the time.

Larger differences in ratings indicate a higher likelihood that the stronger player will win. You should also check that rating changes are reasonable and stable over time, especially for top players.

# Common Questions {.unnumbered}

**Why are ratings not changing much?**  
This may be due to a low K-factor.

**Why do results look incorrect?**  
Ensure matches are in chronological order.

**Can I include ties?**  
Yes, but they must be encoded correctly depending on the package.

# Additional Resources {.unnumbered}

R Packages

- "PlayerRatings": Includes Elo, Glicko, and other rating systems; documentation shows how to prepare and update ratings. [@playerratings]

- "BradleyTerry2": Fits Bradley–Terry paired comparison models; includes examples and vignettes. [@bradleyterry2]

- "elo": Focused specifically on Elo ratings; provides clear examples for calculating ratings from win/loss data. [@elo]
=======
## Step 5: Evaluation {.unnumbered}

After we estimate ratings, we evaluate how well the model performs. This can include checking predictive accuracy on future matches. It could also include comparing results under different K-factors, or examining rating stability over time. Visualizing ratings helps identify inconsistencies.
```
ratings_df <- as.data.frame(elo_ratings$ratings) %>% arrange(desc(Rating))

barplot(ratings_df$Rating,
  names.arg = ratings_df$Player,
  las = 2,
  xlab = "Players",
  ylab = "Elo Rating",
  main = "Elo Ratings of Competitors",
  col = "steelblue")
```

## Step 6: Interpretation {.unnumbered}

Finally, interpret the results of the ratings. Use them to rank the different competitors, estimate win probabilities for future matches, and evaluate uncertainty. Larger differences in Elo ratings indicate a higher likelihood that the stronger player will win. Also consider whether the rating changes are reasonable and stable across matches.

### Additional Resources {.unnumbered}

R Packages

- "PlayerRatings": Includes Elo, Glicko, and other rating systems; documentation shows how to prepare and update ratings.

- "BradleyTerry2": Fits Bradley–Terry paired comparison models; includes examples and vignettes.

- "elo": Focused specifically on Elo ratings; provides clear examples for calculating ratings from win/loss data.
>>>>>>> 1e05f80108bfce3289b835ae9a0ad58ca865960f

Documentation and Vignettes

- Help files in R (e.g., ?PlayerRatings)

- CRAN documentation

- Example workflows showing data preparation and model execution

Research Papers

Original papers on Elo and Glicko, referenced in Glickman and Jones (2024), explain the statistical theory behind the models and how ratings are updated and interpreted [@glickman2024].

<<<<<<< HEAD
# References {.unnumbered}
=======
## References {.unnumbered}
>>>>>>> 1e05f80108bfce3289b835ae9a0ad58ca865960f
